A Test of the Frequency Independence Assumption of Power System Coefficients used in Geomagnetically Induced Current Estimates

### 2019-08-05 paper-submitted-2.tex
Updates from Overleaf

\begin{keypoints}
\item Power system coefficients derived from a time series of GIC and geoelectric field measurements are frequency dependent.
\item A GIC model with frequency-dependent system coefficients provides significantly better estimates of GIC than a frequency independent model.
\item Data-derived power system coefficients may differ significantly with those computed using power system configuration information.
\end{keypoints}

\begin{abstract}
  A common assumption used when estimating geomagnetically induced currents (GICs) in a power system given a time series of nearby direct measurements or indirect estimates of the horizontal geoelectric field components $E_x(t)$ and $E_y(t)$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency-independent power system coefficients.  A first test of this assumption is made using GIC measurements in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away. A second model (Model 2) is obtained using a frequency domain generalization of Model 1: $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$. The coefficients $a(\omega)$ and $b(\omega)$ are shown to be frequency dependent and this model provides significantly better estimates of the measured GIC than Model~1. Based on results using a simulated geoelectric field, it is suggested that the measurement-derived frequency dependence of the system coefficients may be explained by spatial variations in the spectrum of the geoelectric field over the spatial extent of the power system. It is also shown that further improvements over Model~2 can be made using frequency-dependent models with the geomagnetic field as an input.
\end{abstract}

@@

The 1-second-cadence GIC data used span a total of 35 days across 10 intervals that start and end at midnight Japan Standard Time (JST) and are listed in Table~\ref{intervals}. These data have been presented previously in \citet{Watari2009} and \cite{Watari2015}. These time intervals are those for which GIC data was made available with the exception of three days for which there were obvious problems with either the electric field or GIC measurements. The GIC dataset includes 1~s raw data and 1~Hz low-pass-filtered raw measurements and the results presented are for the 1~Hz low-pass-filtered GIC measurements.

GIC was measured in a grounded neutral point of a Y-connected three-phase transformer connected to the 187 kV bus at the Memanbetsu substation of the Hokkaido Power Co. The 187 kV line extends from the Ashro substation to the Memanbetsu substation, which are both grounded. Memanbetsu is an end point of the 187 kV line.

@@

The frequency dependent signal-to-noise ratio for each 1-day segment at a given evaluation frequency is the ratio of of the smoothed power spectrum value of $GIC(\omega_e)$ to the smoothed power spectrum value of the prediction error at that evaluation frequency. A smoothed spectrum was obtained by averaging the raw spectrum in the same frequency bands used for the regression described in Section~\ref{section:Models_and_Methods}. The average signal-to-noise ratio at a given evaluation frequency is the average of the 35 out-of-sample signal-to-noise ratios at that frequency.

@@

\begin{table}
  \caption{Out-of-sample model performance metrics for Model $m$. MSE$_1$/MSE$_m$ is the MSE ratio of Model~1 with respect to that of Model $m$. The PE and MSE ratios are the average of 35 values and the confidence interval, CI, was determined from 1000 bootstrap samples.}
  \centering
  \begin{tabular}{l c c c c}
    \hline
    $m$\hspace{1em} Model & PE & 95\% CI & MSE$_1$/MSE$_m$ & 95\% CI\\
    \hline
    2\hspace{1em} $G_E(t) = a(\omega)E_x(t) + b(\omega)E_y(t)$ & 0.60 & [0.54, 0.65] & 1.8 & [1.6, 2.0]\\
    3\hspace{1em} $G_{E'}(t) = a(\omega)E'_x(t) + b(\omega)E'_y(t)$ & 0.78 & [0.76, 0.80] & 3.1 & [2.8, 3.4]\\
    4\hspace{1em} $G_{B}(t) = z_xB_x(t) + z_yB_y(t)$ & 0.83 & [0.80, 0.85] & 4.6 & [3.8, 5.3]\\
    \hline
  \end{tabular}
  \label{resultstable}
\end{table}

The primary conclusions of this work follow from the results shown in Table~\ref{resultstable}: the frequency dependent model, Model~2, is significantly better at estimating GIC than the frequency independent model, Model~1; Model~3, which uses an estimate of the electric field based on magnetic field measurements, provides significantly better estimates of GIC than Model~2; Model~4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides better estimates of GIC than Model~3.

@@

Figure~\ref{Z_MT} and~\ref{Phi_MT} show the magnitude and phase for the coefficients of $\mathbf{Z}$ in Model~3. Note that transfer function coefficients shown in Figure~\ref{Z_MT} differ significantly from that of a layered slab conductivity model for which $Z_{xy}=-Z_{yx}$ and $Z_{xx}=Z_{yy}=0$. The electric and magnetic field data from Memanbetsu, Japan were also used by \cite{Fujii2015} to compute $\mathbf{Z}$. When using the \cite{Fujii2015} transfer function for Model~3, the PE was significantly lower (0.58 vs. 0.78). Although not critical to the conclusions in this work, a preliminary investigation suggests several explanations. We have found that when using continuous one-month intervals prior to 2006 to compute $\mathbf{Z}$, our calculated transfer function coefficients match the \cite{Fujii2015} coefficients to within the error bars except below periods of 100~s, which is likely explained by the issues with the electric field response in the 1-second-cadence measurements discussed in \cite{Fujii2015} and the fact that we did not use a remote reference in our calculations. However, using one-month intervals after 2006 (January through May tested) our calculated transfer function coefficient magnitudes differ significantly from that in \cite{Fujii2015} - they are shifted downward by approximately 10-50\%. In addition, we have found that the estimated transfer function is sensitive to whether or not segment averaging was used. Due to the large gaps between and the varying length of continuous segments of data used in this work, transfer functions were computed using one-day segments of data and then averaged. We have found that if a continuous span of measurements over 21 days is used to compute a single transfer function, the transfer function coefficient magnitudes are up to 25\% larger than if 21 transfer functions were computed and then averaged. This result is unexpected, and results from extensive verification tests of our MT calculations along with the fact that this effect has observed by both authors using two different MT calculation codes on different data sets (the other code used was LEMI MT V1.4), suggests the explanation is not due to an error in the code used to compute the transfer functions in this work. This result suggests that stack averaging can be a significant source of transfer function bias and we expect that the explanation involves variations in the signal-to-noise between segments.

@@

\cite{Lehtinen1985} justifies the use of the quasi-DC approximation using the argument that GIC amplitudes are significant at frequencies that are much lower than the 50~Hz power line frequency. Clarified, this is essentially a statement that the reactance is proportional to frequency and although the reactance is typically $\sim$5 times larger than the resistance at 50~Hz \citep{Purchala2005}, the reactance will be $\sim$10 times smaller or less than the resistance at GIC frequencies of $\sim$1~Hz and lower. Given these arguments, the quasi-DC approximation is justified, and an explanation involving a violation of the assumption of a uniform half-space conducting slab Earth using a conductivity that varies with depth to more accurately model skin an proximity effects also seems unlikely to explain the results as it would need to account for at least an order of magnitude of reactance. (However, \cite{Boteler1994} noted that for certain systems, the $L/R$ time constant of the power network may be large enough to warrant inclusion in GIC flow calculations; this topic has more recently been pursued by \cite{Oyedokun2013a}.)

@@

Although the use of frequency independent $a$ and $b$ coefficients estimated from data give a reasonable estimates of GIC, it is not known how much data-estimated coefficient values will differ from those directly calculated. (Due to the limited availability of details of the power system and the ground conductivity along the line, we were not able to make this comparison.) Works that use Model~1 and generally note a high degree of visual consistency between the measured values and predictions using this model (e.g., the top panel of Figure~\ref{predictions}). However, we have shown using contemporaneous GIC and electric field measurements that a significantly better estimation of GIC can be obtained by using a frequency dependent model of the form of Models~2 and 3.

>>>>

\begin{keypoints}
\item Power system coefficients derived from a time series of GIC and geoelectric field measurements are frequency dependent.
\item A GIC model with frequency-dependent system coefficients provides significantly better estimates of GIC than a frequency-independent model.
\item Data-derived power system coefficients may differ significantly with those computed using power system configuration information.
\end{keypoints}

\begin{abstract}
  A common assumption used when estimating geomagnetically induced currents (GICs) in a power system given a time series of nearby direct measurements or indirect estimates of the horizontal geoelectric field components $E_x(t)$ and $E_y(t)$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency-independent power system coefficients.  A first test of this assumption is made using GIC measurements in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away. A second model (Model 2) is obtained using a frequency domain generalization of Model 1: $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$. The coefficients $a(\omega)$ and $b(\omega)$ are shown to be frequency-dependent and this model provides significantly better estimates of the measured GIC than Model~1. Based on results using a simulated geoelectric field, it is suggested that the measurement-derived frequency dependence of the system coefficients may be explained by spatial variations in the spectrum of the geoelectric field over the spatial extent of the power system. It is also shown that further improvements over Model~2 can be made using frequency-dependent models with the geomagnetic field as an input.
\end{abstract}

@@

The 1-second-cadence GIC data used span a total of 35 days across 10 intervals that start and end at midnight Japan Standard Time (JST) and are listed in Table~\ref{intervals}. These data have been presented previously in \citet{Watari2009} and \cite{Watari2015}. These time intervals are those for which GIC data was made available except for three days for which there were obvious problems with either the electric field or GIC measurements. The GIC dataset includes 1~s raw data and 1~Hz low-pass-filtered raw measurements and the results presented are for the 1~Hz low-pass-filtered GIC measurements.

GIC was measured in a grounded neutral point of a Y-connected three-phase transformer connected to the 187 kV bus at the Memanbetsu substation of the Hokkaido Power Co. The 187 kV line extends from the Ashro substation to the Memanbetsu substation, which are both grounded. Memanbetsu is an end-point of the 187 kV line.

@@

The frequency-dependent signal-to-noise ratio for each 1-day segment at a given evaluation frequency is the ratio of the smoothed power spectrum value of $GIC(\omega_e)$ to the smoothed power spectrum value of the prediction error at that evaluation frequency. A smoothed spectrum was obtained by averaging the raw spectrum in the same frequency bands used for the regression described in Section~\ref{section:Models_and_Methods}. The average signal-to-noise ratio at a given evaluation frequency is the average of the 35 out-of-sample signal-to-noise ratios at that frequency.

@@

\begin{table}
  \caption{Out-of-sample model performance metrics for Model $m$. MSE$_1$/MSE$_m$ is the MSE ratio of Model~1 to that of Model $m$. The PE and MSE ratios are the average of 35 values and the confidence interval, CI, was determined from 1000 bootstrap samples.}
  \centering
  \begin{tabular}{l c c c c}
    \hline
    $m$\hspace{1em} Model & PE & 95\% CI & MSE$_1$/MSE$_m$ & 95\% CI\\
    \hline
    2\hspace{1em} $G_E(t) = a(\omega)E_x(t) + b(\omega)E_y(t)$ & 0.60 & [0.54, 0.65] & 1.8 & [1.6, 2.0]\\
    3\hspace{1em} $G_{E'}(t) = a(\omega)E'_x(t) + b(\omega)E'_y(t)$ & 0.78 & [0.76, 0.80] & 3.1 & [2.8, 3.4]\\
    4\hspace{1em} $G_{B}(t) = z_xB_x(t) + z_yB_y(t)$ & 0.83 & [0.80, 0.85] & 4.6 & [3.8, 5.3]\\
    \hline
  \end{tabular}
  \label{resultstable}
\end{table}

The primary conclusions of this work follow from the results shown in Table~\ref{resultstable}: the frequency-dependent model, Model~2, is significantly better at estimating GIC than the frequency-independent model, Model~1; Model~3, which uses an estimate of the electric field based on magnetic field measurements, provides significantly better estimates of GIC than Model~2; Model~4, in which a transfer function that connects GIC to $\mathbf{B}$ was derived directly from the data, provides better estimates of GIC than Model~3.

@@

Figure~\ref{Z_MT} and~\ref{Phi_MT} show the magnitude and phase for the coefficients of $\mathbf{Z}$ in Model~3. Note that transfer function coefficients shown in Figure~\ref{Z_MT} differ significantly from that of a layered slab conductivity model for which $Z_{xy}=-Z_{yx}$ and $Z_{xx}=Z_{yy}=0$. The electric and magnetic field data from Memanbetsu, Japan were also used by \cite{Fujii2015} to compute $\mathbf{Z}$. When using the \cite{Fujii2015} transfer function for Model~3, the PE was significantly lower (0.58 vs. 0.78). Although not critical to the conclusions in this work, a preliminary investigation suggests several explanations. We have found that when using continuous one-month intervals prior to 2006 to compute $\mathbf{Z}$, our calculated transfer function coefficients match the \cite{Fujii2015} coefficients to within the error bars except below periods of 100~s, which is likely explained by the issues with the electric field response in the 1-second-cadence measurements discussed in \cite{Fujii2015} and the fact that we did not use a remote reference in our calculations. However, using one-month intervals after 2006 (January through May tested) our calculated transfer function coefficient magnitudes differ significantly from that in \cite{Fujii2015} - they are shifted downward by approximately 10-50\%. Also, we have found that the estimated transfer function is sensitive to whether or not segment averaging was used. Due to the large gaps between and the varying length of continuous segments of data used in this work, transfer functions were computed using one-day segments of data and then averaged. We have found that if a continuous span of measurements over 21 days is used to compute a single transfer function, the transfer function coefficient magnitudes are up to 25\% larger than if 21 transfer functions were computed and then averaged. This result is unexpected, and results from extensive verification tests of our MT calculations along with the fact that this effect has observed by both authors using two different MT calculation codes on different data sets (the other code used was LEMI MT V1.4), suggests the explanation is not due to an error in the code used to compute the transfer functions in this work. This result suggests that stack averaging can be a significant source of transfer function bias and we expect that the explanation involves variations in the signal-to-noise between segments.

@@

\cite{Lehtinen1985} justifies the use of the quasi-DC approximation using the argument that GIC amplitudes are significant at frequencies that are much lower than the 50~Hz power line frequency. Clarified, this is essentially a statement that the reactance is proportional to frequency and although the reactance is typically $\sim$5 times larger than the resistance at 50~Hz \citep{Purchala2005}, the reactance will be $\sim$10 times smaller or less than the resistance at GIC frequencies of $\sim$1~Hz and lower. Given these arguments, the quasi-DC approximation is justified, and an explanation involving a violation of the assumption of a uniform half-space conducting slab Earth using a conductivity that varies with depth to more accurately model skin and proximity effects also seems unlikely to explain the results as it would need to account for at least an order of magnitude of reactance. (However, \cite{Boteler1994} noted that for certain systems, the $L/R$ time constant of the power network may be large enough to warrant inclusion in GIC flow calculations; this topic has more recently been pursued by \cite{Oyedokun2013a}.)

@@

Although the use of frequency-independent $a$ and $b$ coefficients estimated from data give reasonable estimates of GIC, it is not known how much data-estimated coefficient values will differ from those directly calculated. (Due to the limited availability of details of the power system and the ground conductivity along the line, we were not able to make this comparison.) Works that use Model~1 and generally note a high degree of visual consistency between the measured values and predictions using this model (e.g., the top panel of Figure~\ref{predictions}). However, we have shown using contemporaneous GIC and electric field measurements that a significantly better estimation of GIC can be obtained by using a frequency-dependent model of the form of Models~2 and 3.
